---
title: "Plotting Points"
---

```{r setup, include=FALSE}
source("setup.R")
```

Simply having a bunch of points is not enough.  We gotta see it!  Here are some examples to get you started.

## The Data

Here we will use the enigmatic *Araptus attenuata* data set as a raw `data.frame`

```{r}
library(scape)
data("araptus")
summary(araptus)
```

and as a `sf` object

```{r}
library(sf)
araptus_sf <- st_as_sf(araptus,coords = c("Longitude","Latitude"))
st_crs(araptus_sf) <- 3857
araptus_sf
```

See [point construction](points.html) for a discussion on the differences between `data.frame` and `sf` objects.


## Plotting `data.frame` Points

A set of points is useful in some context.   Here is how we can plot these points using the normal ggplot routines. The background `map_data` is an extension of the `ggplot` library to grab map sections.  Here is the polygon for Mexico.

```{r}
library(ggplot2)
library(dplyr)
library(DT)
map_data("world") %>%
  filter( region == "Mexico", is.na(subregion)) -> baja
datatable( baja )
```

which as you can see is a dataframe of points to be connected into a single object.  We can plot it using normal approaches as:

```{r}
ggplot() + 
  geom_polygon( aes(long,lat), fill="grey", data=baja ) + 
  geom_point( aes(Longitude,Latitude), data=araptus ) + 
  xlab("Longitude") + 
  ylab("Latitude") + 
  coord_map()
```


## Plotting `sf` Objects

The `sf` library also has a bunch of built-in plotting routines.  If you plot a `sf` object, it will make a plot for each of the columns of the data in the object.  For example, the `araptus_sf` object has the following data associated with each point

```{r}
names(araptus_sf)
```

and as such will plot one graph for each column.

```{r}
plot(araptus_sf)
```

You can specify a single plot as (and set default values:

```{r}
plot( araptus_sf[4], main="Habitat Suitability", pch=16, axes=TRUE)
```

It is also possible to plot `sf` objects using `ggplot` as they supply appropriate `geom_` and `coord_` objects.

```{r}
ggplot() + 
  geom_sf( data = araptus_sf, aes(size=Suitability) )  +
  theme_void() +
  coord_sf( datum=NA )
```


## Interactive Maps

Since this document is in HTML, we can take advantage of interactive visualizations.  This is a good thing &copy;.  Here is an interactive leaflet map with each sampling locale designated by a circle whose radius is proportional to the habitat suitability of the host plant, all plot on the ESRI World Topo map.

```{r}
library(leaflet)
leaflet(araptus) %>%
  addProviderTiles( providers$Esri.WorldTopoMap) %>%
  addCircleMarkers( radius = ~Suitability*10, color="green",label = ~Site)
```

We can even get a bit more fancy in making pie charts representing the numbrer of males and females in each population, scaled by the habitat suitability.


```{r}
library(leaflet.minicharts)

leaflet() %>%
  addProviderTiles( providers$Esri.WorldTopoMap) %>%
  addMinicharts( 
    araptus$Longitude, araptus$Latitude,
    type = "pie",
    chartdata = araptus[, c("Males","Females")],
    width = araptus$Suitability*30
    )
```


