---
title: "Categorical Rasters"
---

```{r setup, include=FALSE}
source("setup.R")
```

Rasters can contain either continuous (e.g., elevation) or categorical (e.g., habitat type) data---they are only fancy matrices after all.  Here is an example of how to set up a categorical raster with a Raster Attribute Table.


## Example Data

These data are from [github](https://github.com/dyerlab/DLab-Spatial).

```{r}
library(raster)
pines <- raster("data/pines.asc")
pines
```

This is a simple raster with two representations

```{r}
pines[1:20]
```

zeros everywhere there is no pines and the category 2 where the conifer canopy exists.

```{r}
plot(pines)
```

For completeness, I'll add a few more rasters.

```{r}
oaks <- raster("data/oaks.asc")
shrubry <- raster("data/shrubs.asc")
```

## Defining Categories

To begin, lets put these three together^[There will be overlap and we are going to pretend it is not an issue, this is a tutorial after all!].


```{r}
forest <- shrubry
forest[ pines > 0 ] <- 2
forest[ oaks > 0 ] <- 3
hist(values(forest))
```

Here is what the actual data look like.

```{r}
forest[1:40]
```

And it is not a factor:

```{r}
is.factor( forest )
```

To categorize it, we need to both make it a factor:

```{r}
forest <- as.factor( forest )
levels( forest )[[1]]
```

To make it more usable, we can define these categories on the raster using a Raster Attribute Table (RAT)

```{r}
forest <- ratify( forest )
rat <- levels( forest )[[1]]
rat$Landcover <- c("Open","Shrubry", "Pines","Oaks")
levels( forest ) <- rat 
forest
```


## Extracting Data

Now we can play around with the levels a bit.  Here, lets summarize a locale by grabbing a point in the middle of the forest raster^[I know, the coordinate system is whack, but again, this is a tutorial].

```{r}
pt <- st_point( c(304402, 4133856) )
st_as_text( pt )
```


then extract the 

```{r}
vals <- extract( forest, as(pt,"Spatial"), buffer=7 )[[1]]
table( factorValues( forest, vals ) )
```

OK, so now, you have a categorical data raster and can play with it.


